Potential competition in alpine bumble bees is mitigated by imperfect trait-matching and subpartitioning within floral morphotypes

### 2021-04-06 ms_2/ms_2_v2.Rmd
added management covariate to models

author: Douglas B. Sponsler^1^*, Fabrice Requier^2^, Katharina Kallnik^1^, Alice Claßen^1^, A. Fabienne Maihoff^1^, Ingolf Steffan-Dewenter^1^

@@

Species coexistence is constrained in part by competition over shared food sources. Bumble bees are classic model organisms for the study of competition and coexistence, and this topic has taken on new urgency as climate-induced range shifts are expected to reshuffle competitive relationships between bumble bee species, particularly along the climatic gradients of the mountain slopes that host the world's richest bumble bee communities. We surveyed floral abundance and recorded bumble bee floral visitation patterns along a ~1400 m elevation gradient in the German Alps. Using these data, we first ask how floral preferences compare across bumble bee species and to what extent these patterns can be explained by trait-matching between bumble bee tongue length and floral morphotype. We then investigate the patterns of diet overlap between bumble bee species at the nested levels of floral morphotype, floral family, and floral genus, and model diet overlap as a function bumble bee tongue length and elevation. We find that while tongue length is a significant predictor of bumble bees' floral morphotype preferences, species with similar tongue lengths exhibit divergent preferences with respect to certain flora morphotypes. This indicates that factors other than tongue length are involved in flower choice, which might partly explain the puzzling coexistence of bumble bee species with similar tongue lengths. Moreover, genus-level diet overlap was approximately one half that of morphotype-level diet overlap, indicating that even when bumble bees exhibit similar biases at the level of floral morphotype, partly due to the constraints of tongue length, floral diversity enables the mitigation of competition via the subpartitioning within floral morphotypes. This suggests that floral species richness may be as important as overall floral abundance in avoiding the competitive exclusion of bumble bee species. Diet overlap was highest at the low end of our elevation gradient, declining sharply over the transition from valley floor to mid-slope, then leveling off or declining weakly for the remainder of the gradient. If diet overlap is interpreted as an index of potential competition, upslope shifts in bumble bees and flora could generate more competitive conditions at high elevations in cases where floral resources are limiting.

@@

# INTRODUCTION

@@

# METHODS

## Field system

@@

## Data sources

### Floral morphology

@@

### Bumble bee tongue length

@@

## Data processing

@@

## Statistical models

@@

### Model 1: Comparing overall visitation probability across floral morphotypes

@@

### Model 2: Visitation probability as a function of floral morphotype and bumble bee tongue length

@@

### Model 3: Visitation probability as a function of floral morphotype and bumble bee species

@@

### Model 4: Aggregate diet ordination

@@

### Model 5: Niche overlap by level and elevation

@@

### Model 6: Genus-level niche overlap by tongue length and elevation

@@

# RESULTS

## Summary of visitation and floral survey data

@@

## Floral preference (Models 1-3)

@@

## Aggregate diet similarity

@@

## Site-level diet overlap (Models 5 and 6)

@@

We recommend caution, however, in interpreting the p-values of this model, since the pairwise comparison of group differences entails a degree of non-independence.

@@

```{r fig6, echo = FALSE,  fig.cap = paste("Niche overlap by tongue length class difference through elevation. Lines show GAM smooths and shaded bands show 95\\% confidence intervals.")}
knitr::include_graphics("../output/niche_gam_21_plot1.png")
```

# DISCUSSION

## Floral preference is only partly explained by the relationship between bumble bee tongue length and floral morphotype

@@

## The influence of tongue length on floral preference does not carry on to genus-level niche overlap, and floral diversity within morphotypes relaxes potential competition between bumble bee species of similar tongue length

While tongue length was a relatively strong (though imperfect) predictor of bumble bee species' morphotype preferences, it was a weak predictor of diet similarity (at the level of floral genus) across bumble bees species, echoing the findnings of @Durieux2000-uk. Diet ordination produced a primary axis corresponding to tongue length (and associated with affinity for legumes) and a secondary axis describing the dispersion of short-tongued species, which is consistent with studies in Poland [@Goulson2008-oe], France [@Durieux2000-uk], and England [@Goulson2005-nz] that analyzed bumble bee diet by similar methods, suggesting that this may be a general pattern among bumble bee communities, at least within temperate Europe. Nevertheless, the ordination explained less than 50% of total variance, and while short-tongued species separated clearly from long- and medium-tongued species, the latter overlapped extensively with each other. These results were corroborated by our HGAM analysis of niche overlap, where the only significant in overall niche overlap was between short- and long-tongued species. A caveat to these analyses, though, is that our bumble bee community included only 3 long-tongued species (*B. gerstaeckeri*, *B. hortorum*, and *B. mendax*), two of which (*B. gerstaeckeri* and *B. mendax*) were relatively rare. It is possible that clearer patterns of tongue-length-based diet differentiation would be seen in a more balanced community.

Community-mean diet overlap at the level of floral morphotype was consistently nearly twice as high as diet overlap at the level of floral genus. From a numerical perspective, this result is unsurprising simply because each morphotype comprises many genera. From an ecological perspective, though, this pattern suggests another answer to the question of how bumble bee species of similar tongue length can coexists: if tongue length primarily constrains bumble bee foraging at the coarse level of floral morphotype, floral diversity *within* morphotypes creates space for bumble bees with similar tongue lengths to avoid actual dietary overlap, thus mitigating interspecific competition.

## Climate, land use, and the future of bumble bee niche relations

Community mean diet overlap was highest at the lowest sites and declined sharply over the transition from valley floor to mid-slope. If this pattern of niche overlap moves upslope in response to climate change, it might result in increased competition potential at higher elevations, which could threaten high-elevation bumble bee species that currently exist under conditions of relatively low niche overlap. In our system, though, it must be noted that the lowest elevation sites are also the ones most influenced by human land use, and it is possible that the elevation pattern evident in our data is driven by factors other than elevational climate variation.

## Conclusion
Coexistence between bumble bee species with similar tongue lengths hinges on the relationships between species-level floral preferences, trait-matching, and the emergent phenomenon of diet similarity. In this study, we clarify these relationships with respect to bumble bees and wild flowers interacting along a mountain slope elevation gradient. In addition to addressing the question of bumble bee species coexistence, our results suggest how patterns of competition may respond to climate and land use change along mountain slopes, an issue of immediate concern for the conservation of bumble bees and their floral mutualists [@Sirois-Delisle2018-vi; @Soroye2020-zb].

This suggests that diversity within floral morphotypes may as important as morphotype diversity *per se* and overall floral abundance in facilitating the coexistence of bumble bee species, an insight which may have implications for the design of floral enhancements aimed at bumble bee conservation.

>>>>

author: Douglas B. Sponsler^1^*, Katharina Kallnik^1^, Fabrice Requier^2^, Alice Claßen^1^, A. Fabienne Maihoff^1^, Johanna Sieger^3^, Ingolf Steffan-Dewenter^1^

@@

Species coexistence is constrained by competition over shared food sources. Bumble bees are classic model organisms for the study of competition and coexistence, and this topic has taken on new urgency as climate and land use change threaten to reshuffle competitive relationships between bumble bee species along the mountain slopes that host the world's richest bumble bee communities. We surveyed floral abundance and recorded bumble bee floral visitation patterns along a ~1400 m elevation gradient in the German Alps. Using these data, we first ask how floral preferences compare across bumble bee species and to what extent these patterns can be explained by trait-matching between bumble bee tongue length and floral morphotype. We then investigate the patterns of diet overlap between bumble bee species at the nested levels of floral morphotype, floral family, and floral genus, and model diet overlap as a function bumble bee tongue length and elevation. We find that while tongue length is a significant predictor of bumble bees' floral morphotype preferences, there exists both convergence in floral preference across bumble bees with different tongue lengths and divergence of floral preference among bumble bees with similar tongue lengths. This indicates that factors other than tongue length are involved in flower choice, which might partly explain the coexistence of bumble bee species with similar tongue lengths. Moreover, genus-level diet overlap was approximately one half that of morphotype-level diet overlap, indicating that even when bumble bees exhibit similar biases at the level of floral morphotype, partly due to the constraints of tongue length, floral diversity enables the mitigation of competition via the subpartitioning within floral morphotypes. Diet overlap was highest at the low end of our elevation gradient, declining sharply over the transition to mid-slope, then leveling off or declining weakly for the remainder of the gradient. Resource simplification at low elevations combined with warming climates could cause increasing spillover of low-elevation bumble bee species into high-elevation sites, introducing novel competition potential likely to be exacerbated by the reforestation of sub-alpine meadows.

@@

# 1. INTRODUCTION

@@

# 2. METHODS

## 2.1 Field system

@@

## 2.2 Data sources

### 2.2.1 Floral morphology

@@

### 2.2.2 Bumble bee tongue length

@@

## 2.3 Data processing

@@

## 2.4 Statistical models

@@

### 2.4.1 Model 1: Comparing overall visitation probability across floral morphotypes

@@

### 2.4.2 Model 2: Visitation probability as a function of floral morphotype and bumble bee tongue length

@@

### 2.4.3 Model 3: Visitation probability as a function of floral morphotype and bumble bee species

@@

### 2.4.4 Model 4: Aggregate diet ordination

@@

### 2.4.5 Model 5: Niche overlap by level and elevation

@@

### 2.4.6 Model 6: Genus-level niche overlap by tongue length and elevation

@@

# 3. RESULTS

## 3.1 Summary of visitation and floral survey data

@@

## 3.2 Floral preference (Models 1-3)

@@

## 3.3 Aggregate diet similarity

@@

## 3.4 Site-level diet overlap (Models 5 and 6)

@@

The response curve of niche overlap to elevation was significantly different (p < 0.01) from a flat line for *short:short*, *medium:long*, and *long:long*, but not for *short:long*, *short:medium*, or *medium:medium* **(Figure 6A)**. Niche overlap among short-tongued species (*short:short*) declined linearly with elevation, while niche overlap between medium- and long-tongued species had a u-shaped response to elevation, with maxima at the extreme of the elevation gradient. Niche overlap between long-tongued species exhibited a w-shaped response with peaks at low-, mid-, and high-elevation, though this erratic pattern is likely driven by small number of long-tongued species (*B. gerstaeckeri*. *B. hortorum*, and *B. mendax*) available for comparison. Compared to the reference intercept level of *short:long*, mean niche overlap was significantly higher for *medium:long* (p = 0.002) and *short:short* comparisons (p = 0.02) **(Figure 6B)**.  Model 6 explained 43.5% of total deviance (adjusted R^2^ = 0.24). We recommend caution in interpreting the p-values of this model, since the pairwise comparison of differences entails a degree of non-independence not accounted for in the structure of the model.

@@

```{r fig6, echo = FALSE,  fig.cap = paste("Niche overlap by tongue length pair and elevation. Lines show GAM smooths of niche overlap through elevation, and shaded bands show 95\\% confidence intervals (A). Dot-whisker plots show 95\\% confidence intervals of intercepts (i.e. mean niche overlap) for each tongue length pair (B).")}
knitr::include_graphics("../output/niche_gam_21.png")
```

# 4. DISCUSSION

@@

While tongue length was a relatively strong (though imperfect) predictor of bumble bee species' morphotype preferences, it was a weak predictor of diet similarity (at the level of floral genus) across bumble bees species, echoing the findnings of @Durieux2000-uk. Diet ordination produced a primary axis corresponding to tongue length (and associated with affinity for legumes) and a secondary axis describing the dispersion of short-tongued species, which is consistent with studies in Poland [@Goulson2008-oe], France [@Durieux2000-uk], and England [@Goulson2005-nz] that analyzed bumble bee diet by similar methods, suggesting that this may be a general pattern among bumble bee communities, at least within temperate Europe. Nevertheless, the ordination explained less than 50% of total variance, and while short-tongued species separated clearly from long- and medium-tongued species, the latter overlapped extensively with each other. These results were corroborated by our HGAM analysis of niche overlap, where overall niche overlap was lowest between short- and long-tongued species and significantly higher among short-tongued species and between medium- and long-tongued species. A caveat to these analyses, though, is that our bumble bee community included only 3 long-tongued species (*B. gerstaeckeri*, *B. hortorum*, and *B. mendax*), two of which (*B. gerstaeckeri* and *B. mendax*) were relatively rare. It is possible that clearer patterns of tongue-length-based diet differentiation would be seen in a more balanced community.

From the perspective of interspecific competition in bumble bees, it is significant that community-mean diet overlap at the level of floral morphotype was consistently nearly twice as high as diet overlap at the level of floral genus. From a purely numerical perspective, this result is unsurprising simply because each morphotype comprises many genera. From an ecological perspective, though, this pattern suggests another answer to the question of how bumble bee species of similar tongue length can coexists: if tongue length primarily constrains bumble bee foraging at the coarse level of floral morphotype, floral diversity *within* morphotypes creates space for bumble bees with similar tongue lengths to avoid actual dietary overlap, thus mitigating interspecific competition.

Community mean diet overlap was highest at the lowest sites and declined sharply over the transition to mid-slope. If this pattern of niche overlap moves upslope in response to climate change, it might result in increased competition potential at higher elevations, which could threaten high-elevation bumble bee species that currently exist under conditions of relatively low niche overlap. In our system, though, it must be noted that the lowest elevation sites are also the ones most influenced by human land use. The elevated niche overlap we observed at the lowest sites could be interpreted as an indication that the deleterious effects of land use intensification at low elevation have simply outpaced those of agricultural abandonment at high-elevation, and mountain bumble bees populations will initially experience resource pressure from below as reforestation slowly closes high-elevation meadows. If the elevated niche overlap we observed at low elevation is to be interepreted as an indication of resource limitation, though, it must be noted that overall resource volume is stable in our study system up to the tree line ecotone at around 1500 m.s.l. (Sponsler et al., submitted); if bumble bee species are forced to converge on similar diets at low elevation sites, it is more likely because the floral communities at these sites have been *simplified* in terms of taxonomic/morphological diversity or spatiotemporal structure of availability.

<!-- I could, of course, test this. But I don't want to right now. -->

## Conclusion

Coexistence between bumble bee species with similar tongue lengths hinges on the relationships between species-level floral preferences, trait-matching, and the emergent phenomenon of diet similarity. In this study, we clarify these relationships with respect to bumble bees and wild flowers interacting along a mountain slope elevation gradient. Our results indicate that bumble bee coexistence is favored by the fact that the morphological association between tongue length and floral morphotype is a soft constraint on floral preference, and there exists both convergence in floral preference across bumble bees with different tongue lengths and divergence of floral preference among bumble bees with similar tongue lengths. Moreover, we find that niche overlap at the level of floral genus is much lower than at the level of floral morphotype, suggesting that the partitioning of resources *within* floral morphotypes functions to mitigate potential competition. This means that diversity *within* floral morphotypes may be as important as overall floral abundance or morphotype-level diversity in facilitating the coexistence of bumble bee species, an insight with practical implications for floral resource management aimed at bumble bee conservation.

The future of mountain bumble bees --- and, indeed, of mountain ecosystems in general --- involves the systematic pressure of climate warming as well as the twin land use pressures of agricultural intensification at low elevations and agricultural abandonment (with subsequent reforestation) at high elevations [@Iserbyt2015-jc; @Walcher2017-kw; @Sirois-Delisle2018-vi; @Soroye2020-zb; @Suzuki-Ohno2020-jv]. The outcomes of these processes with respect to competition and coexistence between bumble bee species are not yet clear, but our finding that interspecific niche overlap (interpreted as an index of potential competiion) is highest at low elevations suggests that land use intensification may already be influencing bumble bee niche relationships. Resource simplification at low elevations combined with warming climates along mountain slopes could cause increasing spillover of low-elevation bumble bee species into the historical ranges of high elevation species, introducing novel competition potential likely to be exacerbated by the gradual reforestation of abandoned sub-alpine meadows.
